Stuff

### 2014-01-29 paper/Background.tex
Background cleanup

\section{Background}

In order to obtain the execution semantics of a multi-threaded program on a
multi-processor system, one has to understand the memory consistency model of
the underlying system. This is necessary because, as opposed to a direct access
to a single-ported memory, the processors a) have store buffers which reorders
the sequence of memory requests from the processor, for faster access (say by
bypassing a store-value to a subsequent load request), and b) communicate with
the memory through several levels of data caches for faster access. The memory
consistency model describes the semantic behavior of such requests and the
system ensures that the cache hierarchy and the store buffers do not violate
the specified semantics.

Memory consistency models can be partitioned into two sub-parts
\cite{Arvind-memory-model}: a) the ``global'' store-atomicity property for each
location/address and b) the ``local'' reordering restrictions between requests
to different locations within each processor. The store-atomicity property per
address can be informally explained as follows: each location behaves like an
atomic register in which no two writes (stores) to that location can happen
simultaneously, and any read of that location (load) returns the previously
written value in that location. Most memory consistency models require that the
memory subsystem obeys the store-atomicity property. The reordering
restrictions dictated by the memory consistency model pertain to reordering
requests within the same processor (unlike the store-atomicty property which is
global across the entire system), but across multiple addresses. For example,
it dictates whether a store request to an address $x$ can be issued to the
memory sub-system before a load request to a different address $y$, which was
generated by the same processor earlier (\ie load $y$ happened before store $x$
is program order, but the order of issue to memory sub-system is reversed).

The mechanism employed by the memory sub-system to ensure store-atomicity per
location, in the presence of caches, is known as \emph{cache coherence}. In the
absence of caches, since there is exactly one copy of the data for each
address, it is easy to ensure cache coherence. The problem comes in the
presence of shared caches -- multiple copies of the same address can exist in
different caches, and if these locations are independently written, then the
same address can end up having different values in two different caches. Each
location would no longer behave like an atomic register unless some sort of
co-ordination happens between the different caches -- this co-ordination is
summed up as the \emph{cache coherence protocol}.

\subsection{Formal definition of store-atomicity}
In this section we give the formal definition of the store-atomicity property.

The requests issued by the processors to the memory subsystem can be modeled
using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
processor $p$, \ie $\reqFn: (\textit{Processor} * \mathbb{N}) @-> \Request$.
The processor name and the position of the request within the processor
uniquely identifies the request.  Figure \ref{req-resp} shows the data-type
corresponding to a request issued from the processor to the memory subsystem.
A request contains the address (\addrQ) corresponding to the request, and a
type indicating whether the request is a load or a store (\desc). In case of a
store request, it also contains data (\dataQ) supplied by the processor for
that address.

Figure \ref{req-resp} also shows the data-type corresponding to a response
issued from the memory subsystem to the processor.  A response contains a label
(\labelR) to identify the request corresponding to the response. The label
contains a processor name and a position, which uniquely identifies a request.
In the case of a load request, the memory sub-system returns \dataR, the data
returned by the memory subsystem for that address. The memory sub-system also
returns a natural number, \timeR, which is used solely in the definition of
store atomicity, and is not returned in a real system. \timeR{} can be informally
thought to represent the position at which this request has been handled
globally among all the requests issued from all the processors.

@@

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response), (r_1.\labelR).\fst = (r_2.\labelR).\fst @->\\
(r_1.\labelR).\snd < (r2.\labelR).\snd @-> r_1.\timeR \le r_2.\timeR
\end{multline*}
\end{defn}

>>>>

\section{Correctness of our Cache Coherence System}

Before we can discuss our protocol framework and it's verification we
must first establish exactly what it means for a system to be
correct.

Intuitively, the easiest view of shared memory is as a single ported
memory multiplexed amongst the many individual client/processors. In
this view, all operations in the system appear to have a single unique
total ordering.

While simple, it is often useful for performance to permit some
reorderings. For instance, in the context of a processor with local
store buffer, we may wish to allow store values to bypass subsequent
load requests. In general, any potential reordering may be considered
a valid memory consistency model, though they may result in a stronger
verification requirement.

As part of our design exploration for our concrete protocol example,
we may want to change the memory consistency model. To this end, we
are using the notion of store-atomicity to define our memory
model. Most memory consistency models require that the memory
subsystem obeys the store-atomicity property and as such our design
should be modifiable to any of these designs.

A store-atomic memory model can be partitioned into two
parts~\cite{Arvind-memory-model}. First, a ``global'' store-atomicity
property stating that for every read and write access to a particular
location in memory occur in some total order with each operation
seeing the effects of previous operations and not later ones. Seconds
a ``local'' reordering restrictions between requests to different
locations within each processor. The advantage of this approach is
that reordering restrictions pertain to reordering requests locally of
the same processor, while the global store-atomicity property is
constant throughout. Thus changing the memory model should not incur a
reverification of the store-atomicity property. For our concrete
protocol example, we will assume that our memory model is weakly
ordered, in that there are inter-address dependencies between
processors.

Now that we have informally discussed the meaning store atomicity, we
can now provide a formal definition.

%% For example, it dictates whether a store request to an address $x$ can
%% be issued to the memory sub-system before a load request to a
%% different address $y$, which was generated by the same processor
%% earlier (\ie load $y$ happened before store $x$ is program order, but
%% the order of issue to memory sub-system is reversed).

%% The mechanism employed by the memory sub-system to ensure store-atomicity per
%% location, in the presence of caches, is known as \emph{cache coherence}. In the
%% absence of caches, since there is exactly one copy of the data for each
%% address, it is easy to ensure cache coherence. The problem comes in the
%% presence of shared caches -- multiple copies of the same address can exist in
%% different caches, and if these locations are independently written, then the
%% same address can end up having different values in two different caches. Each
%% location would no longer behave like an atomic register unless some sort of
%% co-ordination happens between the different caches -- this co-ordination is
%% summed up as the \emph{cache coherence protocol}.

\subsection{Formal definition of store atomicity}

To properly define the interface with between the processor and the
caching system we must fine the request and responses for each
processor.  Figure~\ref{req-resp} shows the
data-type corresponding to a request issued from the processor to the
memory subsystem.  A request contains the address (\addrQ)
corresponding to the request, and a type indicating whether the
request is a load or a store (\desc). In case of a store request, it
also contains data (\dataQ) supplied by the processor for that
address.A response contains a label
(\labelR) to identify the request corresponding to the response. The label
contains a processor name and a position, which uniquely identifies a request.
In the case of a load request, the memory sub-system returns \dataR, the data
returned by the memory subsystem for that address. The memory sub-system also
returns a natural number, \timeR, which is used solely in the definition of
store atomicity, and is not returned in a real system. \timeR{} can be informally
thought to represent the position at which this request has been handled
globally among all the requests issued from all the processors.

We model request of a processor formally with the function
\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
\Request$.  The processor name and the position of the request within
the processor uniquely identifies a request.

@@

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.

\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response), (r_1.\labelR).\fst = (r_2.\labelR).\fst @->\\
(r_1.\labelR).\snd < (r2.\labelR).\snd @-> r_1.\timeR \le r_2.\timeR
\end{multline*}
\end{defn}

### 2014-01-29 paper/Introduction.tex
Background cleanup

:qh\section{Introduction}
\label{sec:Introduction}

@@

Since the correctness of this abstraction is paramount in the correct
functioning of the system, significant effort is placed in making sure
that the cache coherence engine is correct. In practice, this is done
in two steps. First, a cache coherence protocol is constructed and
verified. This protocol generally directly represents many of
the key concurrent aspects of an implementation, \eg reads and
writes to the system are broken into local atomic micro-operations, but
ignores implementation details \eg buffer sizing and memory access
times. Second, once the protocol has been verified it is handed to
hardware experts who fill in the missing implementation details and
realize it at the Register Transfer Level (RTL) of gates, wires, and
registers. This involves further breaking down of local micro-operations to
make it realizable in hardware. This effort is not a one time translation, as
performance properties are not easily estimated; a significant amount of
performance testing is requires to select a good implementation.

@@

Because verified protocols are often described via Term Rewrite Systems (TRS).
it was shown by Dave \etal{} that a verified protocol could be transliterated
into Bluespec, a hardware description language based on Guarded Atomic Actions,
a restriction of TRS \cite{Hoe:TCAD, DNA:CoherenceImplementation}. This reduces
the correctness of the implementation to the correctness of the protocol,
effectively removing any implementation bugs. Though in the current versions of
the Bluespec language, each atomic action corresponds to one hardware clock
cycle, there has been some work to reduce this restriction \cite{Karczmarek}.
%This requires that the protocol description includes many important
%implementation-level details which had previously been elided for simplicity
%of the description. Practically, beyond some additional descriptional noise in
%the proof this is not a show stopper.

@@

In practice, protocol variants rely on a small set of general ``local building
block'' properties, \eg{} messages always get from source to destination or
that the local cache's reaction to receiving a protocol message results in the
appropriate messages being sent, the appropriate local state being changed \etc{}.
% about parts of the
%design to be true for the full system correctness.
These properties are easily verifiable since they correspond to a small
partition of the system, for instance a single cache.
The details of how these properties hold vary depending on the precise design
variant. For instance, the reason why some property holds, like the
point-to-point FIFO ordering for messages transferred between two nodes,
depends on the type of network; the proof will be very different for a mesh
network compared to a point-to-point network.

In this paper, we describe an approach to provide this proof framework for any
particular cache coherence protocol via the example of a MSI-style protocol and
show how it can be extended to other protocols like MESI and MOESI. We show how
the modular structure of the proof and design can be shared and how
modification of micro-archtectural aspects of design directly translates to
just reverification of our local building block properties. We use Coq for our
proof framework. We also show how an atomic action based MSI protocol
description can be directly transliterated into an Inductive definition in Coq,
and verify its local building block properties formally, thus completing the
picture.

>>>>

\section{Introduction}
\label{sec:Introduction}

@@

Since the correctness of this abstraction is paramount in the correct
functioning of the system, significant effort is placed in making sure
that the cache coherence engine is correct. In practice, this is done
in two steps. First, a cache coherence protocol is constructed and
verified. This protocol generally directly represents many of
the key concurrent aspects of an implementation, \eg{} reads and
writes to the system are broken into local atomic micro-operations, but
ignores implementation details \eg{} buffer sizing and memory access
times. Second, once the protocol has been verified it is handed to
hardware experts who fill in the missing implementation details and
realize it at the Register Transfer Level (RTL) of gates, wires, and
registers. This involves further breaking down of local micro-operations to
make it realizable in hardware. This effort is not a one time translation, as
performance properties are not easily estimated; a significant amount of
performance testing is requires to select a good implementation.

@@

Because verified protocols are often described via Term Rewrite
Systems (TRS).  it was shown by Dave \etal{} that a verified protocol
could be transliterated into Bluespec, a hardware description language
based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD,DNA:CoherenceImplementation}. This reduces the
correctness of the implementation to the correctness of the protocol,
effectively removing any implementation bugs. Though in the current
versions of the Bluespec language, each atomic action corresponds to
one hardware clock cycle, there has been some work to reduce this
restriction~\cite{Karczmarek}.

@@

In practice, protocol variants rely on a small set of general ``local
building block'' properties, \eg{} messages always get from source to
destination or that the local cache's reaction to receiving a protocol
message results in the appropriate messages being sent, the
appropriate local state being changed \etc{}.  These properties are
easily verifiable since they correspond to a small partition of the
system, for instance a single cache.  The details of how these
properties hold vary depending on the precise design variant. For
instance, the reason why some property holds, like the point-to-point
FIFO ordering for messages transferred between two nodes, depends on
the type of network; the proof will be very different for a mesh
network compared to a point-to-point network.

In this paper, we describe an approach to provide this proof framework
for any particular cache coherence protocol via the example of a
MSI-style protocol and show how it can be extended to other protocols
like MESI and MOESI. We show how the modular structure of the proof
and design can be shared and how modification of microarchitectural
aspects of design directly translates to just reverification of our
local building block properties. We use Coq for our proof
framework. We also show how an atomic action based MSI protocol
description can be directly transliterated into an Inductive
definition in Coq, and verify its local building block properties
formally, thus completing the picture.

\noindent\textbf{Paper Outline}. In Section~\ref{Sec:Background} we
discuss our notion of correctness for our cache coherence protocol
framework. We discuss our MSI protocol example and its implementation
in Section~\ref{Sec:Protocol}. In Section~\ref{Sec:Proof} we discuss
the structuring of our proof framework and verification of the base
protocol. We show implementation refinments in
Section~\ref{Sec:Refinements} and conclude with a general discussion
of the applicability of our technique.
